Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology

### 2023-06-18 manuscript/manuscript.qmd
link to OSF and zenodo

@@

Heterogeneity is inflated in heterogeneity estimates of uncorrected effect sizes due to the variability in reliability between primary studies. This is because the attenuation in effect size will then differ from study to study, which results in differences across studies in observed effect sizes beyond their true variability. This is most easily illustrated by assuming a common true effect size $\mu$ across studies (i.e., a fixed effect and no true heterogeneity) and infinite sample size within studies.

 Any observed effect size $\hat{\mu}$ can then be seen as the true effect size $\mu$ multiplied by a study’s reliability $R_i$ (assuming $R_i = R_{xx’} = R_{yy’}$). Hence, the distribution of observed effect sizes is the true effect size $\mu$ multiplied by the distribution of reliabilities across studies. That is, $\hat{\mu} \sim \mu \times N(\bar{R}, \sigma_R) = N(\mu \bar{R}, \mu \sigma_R)$, where $\bar{R}$ is the average reliability across studies and $\sigma_R$ its standard deviation. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that the observed average effect size will be less than the true effect size by a factor $\bar{R}$ and estimated heterogeneity will be larger than true heterogeneity (i.e., zero) by the product of $\mu \times \sigma_R$ (if heterogeneity is expressed as standard deviation rather than variance). That is, variability in reliability across studies is a force that inflates heterogeneity depending on the size of the true effect size. Table 1 illustrates this effect for three studies with differing reliability and zero true heterogeneity.

@@

@@

All code and data for this project are available at (OSF LINK). For our simulations and analyses we used R `r paste(R.Version()$major, R.Version()$minor, sep = ".")` [@rcoreteam2020] and we used the R package 'metafor' [@10.18637/jss.v036.i03] to perform meta-analyses and estimate heterogeneity. For parallel computing we took advantage of the R package 'parabar' [@constantin2023] and for data handling 'data.table' [@dowle2021] and 'ggplot2' [@wickham2016].

@@

```{r Figure 2, echo=FALSE, out.width = '90%'}

#| fig-cap: "Unreliability in primary studies leads to a net negative bias when heterogeneity is positive. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to truncation in Pearson’s r or translation to Fisher’s z these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15, within-study sample size = 150, and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."

@@

```{r Figure 3, echo=FALSE, out.width = '90%'}
#| fig-cap: "Unreliability in primary studies typically leads to a net negative bias even for small degrees of heterogeneity. Effect size is Pearson’s r and rows indicate the number of studies in the meta-analysis. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to effect size truncation in Pearson’s r these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15, within-study sample size = 150. Code to reproduce figure: osf.io/8ygwj."

@@

```{r Figure 4, echo=FALSE, out.width = '90%'}
#| fig-cap: "A smaller sample size in primary studies tends to exacerbate the bias in observed heterogeneity estimates due to unreliability. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Rows indicate the (fixed) sample size within studies and columns indicate nominal true between-studies standard deviation. Due to effect size truncation in Pearson's $r$ these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15 and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."

>>>>

\flushbottom
\thispagestyle{empty}

\vspace*{\fill}

\fbox{\parbox{\textwidth}{%
All code and data for this project are available on the Open Science Framework (OSF) at https://osf.io/tg68 and permanently archived at Zenodo (doi.org/10.5281/zenodo.8052030).
}}

\clearpage

@@

@@

Heterogeneity is inflated in heterogeneity estimates of uncorrected effect sizes due to the variability in reliability between primary studies. This is because the attenuation in effect size will then differ from study to study, which results in differences across studies in observed effect sizes beyond their true variability. This is most easily illustrated by assuming a common true effect size $\mu$ across studies (i.e., a fixed effect and no true heterogeneity) and infinite sample size within studies. Any observed effect size $\hat{\mu}$ can then be seen as the true effect size $\mu$ multiplied by a study’s reliability $R_i$ (assuming $R_i = R_{xx’} = R_{yy’}$). Hence, the distribution of observed effect sizes is the true effect size $\mu$ multiplied by the distribution of reliabilities across studies. That is, $\hat{\mu} \sim \mu \times N(\bar{R}, \sigma_R) = N(\mu \bar{R}, \mu \sigma_R)$, where $\bar{R}$ is the average reliability across studies and $\sigma_R$ its standard deviation. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that the observed average effect size will be less than the true effect size by a factor $\bar{R}$ and estimated heterogeneity will be larger than true heterogeneity (i.e., zero) by the product of $\mu \times \sigma_R$ (if heterogeneity is expressed as standard deviation rather than variance). That is, variability in reliability across studies is a force that inflates heterogeneity depending on the size of the true effect size. Table 1 illustrates this effect for three studies with differing reliability and zero true heterogeneity.

@@

All code and data for this project are available at the Open Science Framework (OSF) at https://osf.io/tg68 and permanently archived at Zenodo (doi.org/10.5281/zenodo.8052030). For our simulations and analyses we used R `r paste(R.Version()$major, R.Version()$minor, sep = ".")` [@rcoreteam2020] and we used the R package 'metafor' [@10.18637/jss.v036.i03] to perform meta-analyses and estimate heterogeneity. For parallel computing we took advantage of the R package 'parabar' [@constantin2023] and for data handling 'data.table' [@dowle2021] and 'ggplot2' [@wickham2016].

@@

```{r Figure 2, echo=FALSE, out.width = '90%'}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Unreliability in primary studies leads to a net negative bias when heterogeneity is positive. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to truncation in Pearson’s r or translation to Fisher’s z these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15, within-study sample size = 150, and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."

@@

```{r Figure 3, echo=FALSE, out.width = '90%'}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Unreliability in primary studies typically leads to a net negative bias even for small degrees of heterogeneity. Effect size is Pearson’s r and rows indicate the number of studies in the meta-analysis. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to effect size truncation in Pearson’s r these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15, within-study sample size = 150. Code to reproduce figure: osf.io/8ygwj."

@@

```{r Figure 4, echo=FALSE, out.width = '90%'}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "A smaller sample size in primary studies tends to exacerbate the bias in observed heterogeneity estimates due to unreliability. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Rows indicate the (fixed) sample size within studies and columns indicate nominal true between-studies standard deviation. Due to effect size truncation in Pearson's $r$ these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15 and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."
